---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data and load libraries

```{r}
library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(corrplot)
library(tictoc)
library(vip)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

```{r}
# Check Names
names(train_data)

# Glimpse the training data
glimpse(train_data) # we will have to remove this characters Aggregate_HH_INC_ACS_13_17, avg_Agg_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17

# structure 
str(train_data)

# missingness
tibble(variable = names(train_data),
       pct_missing = colMeans(is.na(train_data)) * 100) |>
  arrange(desc(pct_missing))
```

### Outcome Variable

```{r}
# Basic Distribution of non_return_rate 
summary(train_data$non_return_rate)

# Non Missing of non_return_rate
skim(train_data$non_return_rate)

# Examine the distribution of the outcome variable (non_return_rate)
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram() +
  theme_minimal()

# State Frequency
train_data |>
  group_by(State_name) |>
  summarize(average_non = mean(non_return_rate),
            count = n()) |>
  arrange(desc(average_non))
```

### Look for variables highly correlated with outcome variable

```{r}
# Get correlation of numeric predictors with outcome variable
correlations <- train_data |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

# Extract just the correlations with the outcome
outcome_cors <- correlations[, "non_return_rate"]

# Sort in order of highest correlations
outcome_cors_sorted <- sort(abs(outcome_cors), decreasing = TRUE)

# Keep only correlations of 0.5 or higher
outcome_cors_filtered <- outcome_cors_sorted[outcome_cors_sorted >= 0.5]

outcome_cors_filtered
```

\*Note to Miguel: It seems like low_response score and self_response_rate could be really similar to our outcome variable. Should we remove them to prevent data leakage?

### Visualize the correlations

```{r}
train_data |>
  ggplot(aes(x = non_return_rate, y = Low_Response_Score)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

train_data |>
  ggplot(aes(x = non_return_rate, y = Renter_Occp_HU_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

train_data |>
  ggplot(aes(x = non_return_rate, y = Self_Response_Rate_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()
```

```{r}
corrplot(train_data)
```

\*Miguel: I got an error with the corrplot function

## Set up cross-validation and parallelization

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

```{r}
# Set up parallelization
library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

# Predictive Modeling

## Model 1: Linear regression (parametric)

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score, Aggregate_HH_INC_ACS_13_17, 
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic fact 
  step_impute_median(all_numeric_predictors()) 

# Mode
model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()
```

\*Miguel - why did you remove Aggregate_HH_INC_ACS_13_17

```{r}
# Variable importance
model0_final <- model0_wf |>
  last_fit(split) 

model0_final |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 
```

### Elastic Net

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score, Aggregate_HH_INC_ACS_13_17, 
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17) |> # remove the problematic fact 
  step_impute_median(all_numeric_predictors()) 


# Mode
model1_mode <- linear_reg(penalty = tune(), mixture = tune()) |> # works best for many variables 
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet") 

# Workflow
model1_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model1_mode)

# resamples
model1_grid <- grid_regular(
  penalty(), 
  mixture(),
  levels = 20
)

model1_tuning <- tune_grid(
  model1_wf,
  resamples = folds,
  grid = model1_grid
)

show_best(model1_tuning)
```

## Model 2: Non-parametric

## Model 3: Random forest (ensembling)

For our third model, we have selected random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predictors considered at each split using the latin hypercube tuning grid to save computation time. For *min_n*, we tune over a range of 5 to 40 data points required in a node before further splitting. For *mtry,* we tune over a range of 5 to 50 predictors to be considered at each split. During tuning, we use 100 trees. The number of trees will be increased to between 200-500 for final model fitting. We also use parallelization to reduce computation time.

In the recipe, several variables are removed, as well as variables with low variance and high correlation with each other. These steps reduce redundancy and also save computation time. We also use 3-fold cross-validation instead of 5-fold cross-validation.

```{r}
# 3 fold cross validation
threefold <- vfold_cv(train_data, v = 3)
```

```{r}
set.seed(20260412)
tic()

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  add_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, # remove the problematic fact 
          Low_Response_Score,
          Self_Response_Rate_ACS_13_17) |> 
  step_impute_median(all_numeric_predictors()) |>
  step_nzv(all_predictors()) |> # remove predictors with low variance
  step_corr(allpredictors(), threshold = 0.9) # remove predictors 90% correlated with each other

# Model
model3_mod <- rand_forest(min_n = tune(), # hyperparameter tuning
                          mtry = tune(),
                          trees = 100) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) # parallelization

# Workflow
model3_wf <- workflow() |>
  add_recipe(model3_rec) |>
  add_model(model3_mod)

# Tuning grid
model3_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 20
)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = threefold,
  grid = model3_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model3_resamples)

# Plot
autoplot(model3_resamples)
toc()
```
