---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data

```{r}
library(tidyverse)
library(here)
library(tidymodels)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

```{r}
# Glimpse the training data
glimpse(train_data)

# Examine the distribution of the outcome variable 
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram()
```

## Set up cross-validation

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

\*Note to Miguel: when I tried to run the random forest model with 5 fold cross validation without parallelization, it took a very long time to run - I ended up quitting and restarting to see if we can make changes to either the number of folds, adding parallelization, or making some other change to save computation time.

# Predictive Modeling

## Model 1: Parametric

## Model 2: Non-parametric

## Model 3: Random forest (ensembling)

For our third model, we have selected random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predcitors considered at each split. We use 200 trees.

```{r}
set.seed(20260412)

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  add_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"))

# Model
model3_mod <- rand_forest(min_n = tune(),
                          mtry = tune(),
                          trees = 200) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = 4)

# Workflow
model3_wf <- workflow() |>
  add_recipe(model3_rec) |>
  add_model(model3_mod)

# Tuning grid 
model3_grid <- grid_regular(
  mtry(range = c(1, 15)),
  min_n(range = c(1, 15)),
  levels = 5
)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = folds,
  grid = model3_grid
)

# Collect metrics
collect_metrics(model3_resamples)

# Plot
autoplot(model3_resamples)
```
