---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data

```{r message=FALSE, warning=FALSE}
options(scipen = 999)

library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(janitor)
library(corrplot)
library(tictoc)
library(vip)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

### Basic Overview

```{r}
# Check Names
names(train_data)

# Glimpse the training data
glimpse(train_data) # we will have to remove this characters Aggregate_HH_INC_ACS_13_17, avg_Agg_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17

# structure 
str(train_data)

# missingness
tibble(variable = names(train_data),
       pct_missing = colMeans(is.na(train_data)) * 100) |>
  arrange(desc(pct_missing))
```

### Outcome Variable

```{r}
# Basic Distribution of non_return_rate
summary(train_data$non_return_rate)

# Non Missing of non_return_rate
skim(train_data$non_return_rate)

# Examine the distribution of the outcome variable (non_return_rate)
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram() +
  labs(title = "Distribution of Non Return Rate") +
  theme_minimal()

train_data |> 
  skewness(non_return_rate)
# State Frequency
train_data |>
  group_by(State_name) |>
  summarize(average_non = mean(non_return_rate),
            count = n(),
            pct_obs = count / nrow(train_data) * 100) |>
  arrange(pct_obs)
```

### Transformations
```{r}
train_data |>
  mutate(
    raw_non   = non_return_rate,
    log_non = log(non_return_rate + 0.001),
    logit_non = log(non_return_rate/ (1 - non_return_rate + 0.001))) |>
  pivot_longer(c(raw_non, log_non, logit_non)) |>
  ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~name, scales = "free") +
    labs(title = "Non Return Rate Transformations") +
    theme_minimal()

```
Will explore log transformation of non_return_rate. There are also other variables, in particular the income, home value, and population variables which could benefit from a log. 

These are
Population: Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17, Pop_under_5_CEN_2010, Pop_18_24_CEN_2010
Value: Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17
Otherss: pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17,
pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17,
pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010,
pct_No_Plumb_ACS_13_17
### Correlations
Target Correlation: 
Typically you better features (non-target variables) that are highly correlated with target variables as they provide more predictive power.
```{r}
numeric_vars <- train_data |>
  select(where(is.numeric), -non_return_rate)

# find correlations
correlations <- cor(numeric_vars, train_data$non_return_rate, use = "complete.obs") 
  
# Convert to dataframe and sort
non_cor <- data_frame(
  correlation = (correlations)) |>
  arrange(desc(abs(correlation)))

non_cor

```

Feature Correlation:
One potential issue can be when features (non-target variables) are highly correlated with one another. As one example percentage of families that speak tagalog and spanish may both contribute similar information to our model, and inclusion of both can lead to overfitting and make our estimates less stable. This issue doesn't matter as much for regularized models but does for regular regression. 
```{r}
corr_mat <- cor(numeric_vars, use = "pairwise.complete.obs")

# convert to long form
corr_data <- as.data.frame(as.table(corr_mat))

# remove self-correlations and duplicates
corr_df <- corr_df |>
  filter(Var1 != Var2) |>
  mutate(abs_corr = abs(Freq)) |>
  arrange(desc(abs_corr))

# top redundant pairs
head(corr_df, 100)
```

Way too many correlated features to even begin to count. I will approach this issue by creating clusters and picking representative variables by cluster. 





PCA can help us look for clustering of features that introduce redundancy into prediction. Basically it'll reduce redundant variables into one component. All i have to add is: 

step_normalize(all_numeric_predictors()) |>
step_pca(all_numeric_predictors(), num_comp = tune())

### Potentially Highly Correlated Relationships

## Set up cross-validation

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

```{r}
library(doParallel)
cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

\*Note to Miguel: when I tried to run the random forest model with 5 fold cross validation without parallelization, it took a very long time to run - I ended up quitting and restarting to see if we can make changes to either the number of folds, adding parallelization, or making some other change to save computation time.

# Predictive Modeling

## Model 1: Parametric

### Plain Vanilla Linear Regression
```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic vars (chr) 
  step_impute_median(all_numeric_predictors()) 

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()

# Variable importance
model0_final <- model0_wf |>
  last_fit(split) 

model0_final |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 

## Extracting the most important for later use 
fit <- model0_final |> 
  extract_fit_parsnip()

imp_tbl <- vi(fit)

# sort by absolute importance
imp_tbl <- imp_tbl |> arrange(desc(abs(Importance)))

# convert to a vector of variable names
top_vars <- imp_tbl$Variable[1:20]

top_vars

top_vars_clean <- intersect(top_vars, names(train_data))
```
```

### With transformations
```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic vars (chr) 
  step_log(non_return_rate, offset = 1e-6) |>
  step_log(Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17,
           Pop_under_5_CEN_2010, Pop_18_24_CEN_2010, offset = 1) |>
  step_log(Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17, offset = 1) |>
  step_log(pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17,
           pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17,
           pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010,
           pct_No_Plumb_ACS_13_17, offset = 1e-6) |>
  step_impute_median(all_numeric_predictors()) |>
  step_zv(all_numeric_predictors()) |> 
  step_normalize(all_numeric_predictors())

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()

```

### Linear Regression with PCA, given the large number of vars. 
```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic vars (chr) 
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = tune())

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

# adding in pca grid 
pca_grid <- grid_regular(
  num_comp(range = c(50, 150)),
  levels = 10)

model0_resamples <- tune_grid(
  model0_wf,
  resamples = folds,
  grid = pca_grid)

model0_resamples |>
  collect_metrics()



```

### Elastic Net

```{r}
set.seed(20260412)

# Mode
model1_mode <- linear_reg(penalty = tune(), mixture = tune()) |> # works best for many variables 
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet") 

# Workflow
model1_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model1_mode)

# resamples
model1_grid <- grid_regular(
  penalty(), 
  mixture(),
  levels = 20
)

model1_tuning <- tune_grid(
  model1_wf,
  resamples = folds,
  grid = model1_grid
)

show_best(model1_tuning)

# Finalize workflow
model1_best <- select_best(model1_tuning, metric = "rmse")


model1_final_wf <- finalize_workflow(
  model1_wf,
  model1_best)

# Variable importance
model1_fit <- fit(model1_final_wf, data = train_data)

model1_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 
```

## Model 2: Non-parametric

## Model 3: Random forest (ensembling)

For our third model, we have selected random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predcitors considered at each split. We use 200 trees.
```{r}
set.seed(20260412)
tic()

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data1) |>
  update_role(State_name, County_name, Low_Response_Score, new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010,
          Aggr_House_Value_ACS_13_17,
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17,
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |>
  step_impute_median(all_numeric_predictors()) |>
  step_select(top_vars)

# Model
model3_mod <- rand_forest(min_n = tune(),
                          mtry = tune(),
                          trees = 100) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) # parallelization

# Workflow
model3_wf <- workflow() |>
  add_recipe(model3_rec) |>
  add_model(model3_mod)

# Tuning grid 
model3_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 12)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = folds,
  grid = model3_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model3_resamples)
```

Model 2 (with only the best fits from earlier )
```{r}
set.seed(20260412)
tic()
train_data1 <- train_data |>
  select(non_return_rate, all_of(top_vars_clean)) 

# Recipe
model4_rec <- recipe(non_return_rate ~ ., data = train_data1) |>
  step_impute_median(all_numeric_predictors()) 

prep(model4_rec) |> juice() |> names()

# Model
model4_mod <- rand_forest(min_n = tune(),
                          mtry = tune(),
                          trees = 100) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) # parallelization

# Workflow
model4_wf <- workflow() |>
  add_recipe(model4_rec) |>
  add_model(model4_mod)

# Tuning grid 
model4_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 20)

# Resamples
model4_resamples <- tune_grid(
  model4_wf,
  resamples = folds,
  grid = model4_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model4_resamples)

```
