---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data and load libraries

```{r message=FALSE, warning=FALSE}
options(scipen = 999)

library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(janitor)
library(corrplot)
library(tictoc)
library(vip)
library(e1071)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

### Basic Overview

```{r}
# Check Names
names(train_data)

# Glimpse the training data
glimpse(train_data) # we will have to remove this characters Aggregate_HH_INC_ACS_13_17, avg_Agg_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17

# structure 
str(train_data)

# missingness
tibble(variable = names(train_data),
       pct_missing = colMeans(is.na(train_data)) * 100) |>
  arrange(desc(pct_missing))
```

### Outcome Variable

```{r}
# Basic Distribution of non_return_rate 
summary(train_data$non_return_rate)

# Non Missing of non_return_rate
skim(train_data$non_return_rate)

# Examine the distribution of the outcome variable (non_return_rate)
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram() +
  theme_minimal() +
  labs(title = "Distribution of Non Return Rate") +
  theme_minimal()

# Examine the skewness of the outcome variable
train_data |> 
  skewness(non_return_rate)
# Note to Miguel: the skewness command is giving me an error

# State Frequency
train_data |>
  group_by(State_name) |>
  summarize(average_non = mean(non_return_rate),
            count = n()) |>
  arrange(desc(average_non))
```

### Look for variables highly correlated with outcome variable

```{r}
# Get correlation of numeric predictors with outcome variable
correlations <- train_data |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

# Extract just the correlations with the outcome
outcome_cors <- correlations[, "non_return_rate"]

# Sort in order of highest correlations
outcome_cors_sorted <- sort(abs(outcome_cors), decreasing = TRUE)

# Keep only correlations of 0.5 or higher
outcome_cors_filtered <- outcome_cors_sorted[outcome_cors_sorted >= 0.5]

outcome_cors_filtered
```

\*Note to Miguel: It seems like low_response score and self_response_rate could be really similar to our outcome variable. Should we remove them to prevent data leakage?

### Visualize the correlations

```{r}
train_data |>
  ggplot(aes(x = non_return_rate, y = Low_Response_Score)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

train_data |>
  ggplot(aes(x = non_return_rate, y = Renter_Occp_HU_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

train_data |>
  ggplot(aes(x = non_return_rate, y = Self_Response_Rate_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()
```

```{r}
train_data |>
  select(where(is.numeric)) |>
  corrplot()
```

\*Miguel: I got an error with the corrplot function

### Transformations

```{r}
train_data |>
  mutate(
    raw_non   = non_return_rate,
    log_non = log(non_return_rate + 0.001),
    logit_non = log(non_return_rate/ (1 - non_return_rate + 0.001))) |>
  pivot_longer(c(raw_non, log_non, logit_non)) |>
  ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~name, scales = "free") +
    labs(title = "Non Return Rate Transformations") +
    theme_minimal()
```

Will explore log transformation of non_return_rate. There are also other variables, in particular the income, home value, and population variables which could benefit from a log.

These are Population: Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17, Pop_under_5_CEN_2010, Pop_18_24_CEN_2010 Value: Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17 Otherss: pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17, pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17, pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010, pct_No_Plumb_ACS_13_17

### Correlations

Target Correlation: Typically you better features (non-target variables) that are highly correlated with target variables as they provide more predictive power.

```{r}
numeric_vars <- train_data |>
  select(where(is.numeric), -non_return_rate)

# find correlations
correlations <- cor(numeric_vars, train_data$non_return_rate, use = "complete.obs") 
  
# Convert to dataframe and sort
non_cor <- data_frame(
  variable = rownames(correlations),
  correlation = (correlations)) |>
  arrange(desc(abs(correlation)))

non_cor

```

One potential issue is a lot of these variables measure the same thing just in different survey (pct_NH_White_alone_CEN_2010, pct_NH_White_alone_ACS_13_17)

Feature Correlation: One potential issue can be when features (non-target variables) are highly correlated with one another. As one example percentage of families that speak tagalog and spanish may both contribute similar information to our model, and inclusion of both can lead to overfitting and make our estimates less stable. This issue doesn't matter as much for regularized models but does for regular regression.

```{r}
corr_mat <- cor(numeric_vars, use = "pairwise.complete.obs")

# distance based
dist_mat <- as.dist(1 - abs(corr_mat))

# hierarchical clustering
hc <- hclust(dist_mat, method = "average")

plot(hc, labels = FALSE, main = "Correlation Clusters")

# Id'ing cluster
clusters <- cutree(hc, h = 0.15)   # 1 - .85
table(clusters)
```

Way too many correlated features to even begin to count. I will approach this issue by creating clusters and picking representative variables by cluster. Cluster 7 seems like its the biggest issue but there are a few blocks of collinearity

```{r}
# Now ill attach cluster numbers to each variable
cluster_df <- data.frame(
  variable = names(numeric_vars),
  cluster  = clusters)

representatives <- cluster_df |>
  group_by(cluster) |>
  slice(1) |> 
  pull(variable)

vars_to_drop <- setdiff(cluster_df$variable, representatives)

vars_to_drop

```

PCA can help us look for clustering of features that introduce redundancy into prediction. Basically it'll reduce redundant variables into one component. All i have to add is:

step_normalize(all_numeric_predictors()) \|\> step_pca(all_numeric_predictors(), num_comp = tune())

## Set up cross-validation

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

```{r}
# Set up parallelization
library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

\*Note to Miguel: when I tried to run the random forest model with 5 fold cross validation without parallelization, it took a very long time to run - I ended up quitting and restarting to see if we can make changes to either the number of folds, adding parallelization, or making some other change to save computation time.

# Predictive Modeling

## Model 1: Parametric

### Plain Vanilla Linear Regression

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17,
          pct_TEA_Update_Leave_CEN_2010) |> # high missingness
  step_impute_median(all_numeric_predictors())

prep(model1_rec) |> juice() |> names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()

```

\*Miguel - why did you remove Aggregate_HH_INC_ACS_13_17

```{r}
# Variable importance
model0_final <- model0_wf |>
  last_fit(split) 

model0_final |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 

## Extracting the most important for later use 
fit <- model0_final |> 
  extract_fit_parsnip()

imp_tbl <- vi(fit)

# sort by absolute importance
imp_tbl <- imp_tbl |> arrange(desc(abs(Importance)))

# convert to a vector of variable names
top_vars <- imp_tbl$Variable[1:20]

top_vars

top_vars_clean <- intersect(top_vars, names(train_data))
```

### Trying to remove redundant features

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # removing census duplicated vars
  step_rm(all_of(vars_to_drop)) |> # removing highly correlated vars
  step_impute_median(all_numeric_predictors())

prep(model1_rec) |> juice() |> names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()
```

Didn't really do anything lol

### With transformations

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic vars (chr) 
  step_log(Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17, offset = 1) |>
  step_log(Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17, offset = 1) |>
  step_log(pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17,
           pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17,
           pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010,
           pct_No_Plumb_ACS_13_17, offset = 1e-6) |>
  step_rm(all_of(vars_to_drop)) |> # removing highly correlated vars
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) 

prep(model1_rec) |> juice() |> names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()

model0_final <- model0_wf |>
  last_fit(split) |>
  collect_metrics()



```

### Linear Regression with PCA, given the large number of vars.

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score,
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          Self_Response_Rate_ACS_13_17) |> # remove the problematic vars (chr) 
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = tune())

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

# adding in pca grid 
pca_grid <- grid_regular(
  num_comp(range = c(50, 150)),
  levels = 10)

model0_resamples <- tune_grid(
  model0_wf,
  resamples = folds,
  grid = pca_grid)

model0_resamples |>
  collect_metrics()

```

### Elastic Net

```{r}
set.seed(20260412)

# Mode
model1_mode <- linear_reg(penalty = tune(), mixture = tune()) |> # works best for many variables 
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet") 

# Workflow
model1_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model1_mode)

# resamples
model1_grid <- grid_regular(
  penalty(), 
  mixture(),
  levels = 20
)

model1_tuning <- tune_grid(
  model1_wf,
  resamples = folds,
  grid = model1_grid
)

show_best(model1_tuning)

# Finalize workflow
model1_best <- select_best(model1_tuning, metric = "rmse")


model1_final_wf <- finalize_workflow(
  model1_wf,
  model1_best)

# Variable importance
model1_fit <- fit(model1_final_wf, data = train_data)

model1_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 
```

## Model 2: Non-parametric

## Model 3: Random forest (ensembling)

For our third model, we have selected random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predictors considered at each split using the latin hypercube tuning grid to save computation time. For *min_n*, we tune over a range of 5 to 40 data points required in a node before further splitting. For *mtry,* we tune over a range of 5 to 50 predictors to be considered at each split. During tuning, we use 100 trees. The number of trees will be increased to between 200-500 for final model fitting. We also use parallelization to reduce computation time.

In the recipe, several variables are removed, as well as variables with low variance and high correlation with each other. These steps reduce redundancy and also save computation time. We also use 3-fold cross-validation instead of 5-fold cross-validation.

```{r}
# 3 fold cross validation
threefold <- vfold_cv(train_data, v = 3)
```

### RF Model 1

```{r}
set.seed(20260412)
tic()

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  add_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, # remove the problematic fact 
          Low_Response_Score,
          Self_Response_Rate_ACS_13_17) |> 
  step_impute_median(all_numeric_predictors()) |>
  step_nzv(all_predictors()) |> # remove predictors with low variance
  step_corr(allpredictors(), threshold = 0.9) |> # remove predictors 90% correlated with each other
  step_select(top_vars)

# Model
model3_mod <- rand_forest(min_n = tune(), # hyperparameter tuning
                          mtry = tune(),
                          trees = 100) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) # parallelization

# Workflow
model3_wf <- workflow() |>
  add_recipe(model3_rec) |>
  add_model(model3_mod)

# Tuning grid
model3_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 20
)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = threefold,
  grid = model3_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model3_resamples)

toc()
```

### RF Model 2 (with only the best fits from earlier )

```{r}
set.seed(20260412)
tic()
train_data1 <- train_data |>
  select(non_return_rate, all_of(top_vars_clean)) 

# Recipe
model4_rec <- recipe(non_return_rate ~ ., data = train_data1) |>
  step_impute_median(all_numeric_predictors()) 

# Model
model4_mod <- rand_forest(min_n = tune(),
                          mtry = tune(),
                          trees = 100) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) # parallelization

# Workflow
model4_wf <- workflow() |>
  add_recipe(model4_rec) |>
  add_model(model4_mod)

# Tuning grid 
model4_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 20)

# Resamples
model4_resamples <- tune_grid(
  model4_wf,
  resamples = folds,
  grid = model4_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model4_resamples)

# Plot
autoplot(model3_resamples)
toc()
```
