---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data

```{r}
library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(janitor)
library(corrplot)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

### Basic Overview

```{r}
# Check Names
names(train_data)

# Glimpse the training data
glimpse(train_data) # we will have to remove this characters Aggregate_HH_INC_ACS_13_17, avg_Agg_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17

# structure 
str(train_data)

# missingness
tibble(variable = names(train_data),
       pct_missing = colMeans(is.na(train_data)) * 100) |>
  arrange(desc(pct_missing))
```

### Outcome Variable

```{r}
# Basic Distribution of non_return_rate
summary(train_data$non_return_rate)

# Non Missing of non_return_rate
skim(train_data$non_return_rate)

# Examine the distribution of the outcome variable (non_return_rate)
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram() +
  theme_minimal()

# State Frequency
train_data |>
  group_by(State_name) |>
  summarize(average_non = mean(non_return_rate),
            count = n()) |>
  arrange(desc(average_non))
```

### Potentially Highly Correlated Relationships

```{r}
train_data |>
  ggplot(aes(x = non_return_rate, y = Low_Response_Score)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

train_data |>
  ggplot(aes(x = non_return_rate, y = Renter_Occp_HU_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()
```

### Correlations

```{r}
corrplot(train_data)

```

## Set up cross-validation

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

```{r}
library(doParallel)
cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

\*Note to Miguel: when I tried to run the random forest model with 5 fold cross validation without parallelization, it took a very long time to run - I ended up quitting and restarting to see if we can make changes to either the number of folds, adding parallelization, or making some other change to save computation time.

# Predictive Modeling

## Model 1: Parametric

### Plain Vanilla Linear Regression

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score, Aggregate_HH_INC_ACS_13_17, 
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17) |> # remove the problematic fact 
  step_impute_median(all_numeric_predictors()) 

# Mode
model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()
```

### Elastic Net

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  update_role(State_name, County_name, Low_Response_Score, Aggregate_HH_INC_ACS_13_17, 
              new_role = "id") |>
  step_rm(has_role("id"),
          pct_TEA_Update_Leave_CEN_2010, # remove the high missingness
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17) |> # remove the problematic fact 
  step_impute_median(all_numeric_predictors()) 


# Mode
model1_mode <- linear_reg(penalty = tune(), mixture = tune()) |> # works best for many variables 
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet") 

# Workflow
model1_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model1_mode)

# resamples
model1_grid <- grid_regular(
  penalty(), 
  mixture(),
  levels = 20
)

model1_tuning <- tune_grid(
  model1_wf,
  resamples = folds,
  grid = model1_grid
)

show_best(model1_tuning)
```

## Model 2: Non-parametric

## Model 3: Random forest (ensembling)

For our third model, we have selected random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predcitors considered at each split. We use 200 trees.

```{r}
set.seed(20260412)

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  add_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"))

# Model
model3_mod <- rand_forest(min_n = tune(),
                          mtry = tune(),
                          trees = 200) |>
  set_mode(mode = "regression") |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = 4)

# Workflow
model3_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model3_mod)

# Tuning grid 
model3_grid <- grid_regular(
  mtry(range = c(1, 15)),
  min_n(range = c(1, 15)),
  levels = 5
)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = folds,
  grid = model3_grid
)

# Collect metrics
collect_metrics(model3_resamples)

# Plot
autoplot(model3_resamples)
```
