---
title: "assignment07"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
---

# Set up

Read in the data and load libraries

```{r message=FALSE, warning=FALSE}
options(scipen = 999)

library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(janitor)
library(corrplot)
library(tictoc)
library(vip)
library(e1071)
library(glmnet)

data <- read.csv(here("data/assignment_modeling.csv"))
```

## Split the data

```{r}
set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)
```

## EDA on the training data

### Basic Overview

```{r}
# Check Names
names(train_data)

# Glimpse the training data
glimpse(train_data) # we will have to remove this characters Aggregate_HH_INC_ACS_13_17, avg_Agg_HH_INC_ACS_13_17, Aggr_House_Value_ACS_13_17

# Check the structure 
str(train_data)

# Examine missingness
tibble(variable = names(train_data),
       pct_missing = colMeans(is.na(train_data)) * 100) |>
  arrange(desc(pct_missing))
```

### Outcome Variable

```{r}
# Basic Distribution of non_return_rate 
summary(train_data$non_return_rate)

# Non Missing of non_return_rate
skim(train_data$non_return_rate)

# Examine the distribution of non_return_rate
train_data |>
  ggplot(aes(x = non_return_rate)) +
  geom_histogram() +
  theme_minimal() +
  labs(title = "Distribution of Non Return Rate") +
  theme_minimal()

# State Frequency of non return rate 
train_data |>
  group_by(State_name) |>
  summarize(average_non = mean(non_return_rate),
            count = n()) |>
  arrange(desc(average_non))
```

### Look for variables highly correlated with outcome variable

```{r}
# Get correlation of numeric predictors with outcome variable
correlations <- train_data |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs")

# Extract just the correlations with the outcome
outcome_cors <- correlations[, "non_return_rate"]

# Sort in order of highest correlations
outcome_cors_sorted <- sort(abs(outcome_cors), decreasing = TRUE)

# Keep only correlations of 0.5 or higher
outcome_cors_filtered <- outcome_cors_sorted[outcome_cors_sorted >= 0.5]

outcome_cors_filtered
```

Low_response_score and self_response_rate are very similar to our outcome variable. We will remove them to prevent data leakage. Many of the variables that are correlated with the outcome variable are related to the breakdown of renters and owners in an area, and the demographic breakdowns.

### Visualize the correlations

```{r}
# Percent owner occupied 
train_data |>
  ggplot(aes(x = non_return_rate, y = pct_Owner_Occp_HU_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

# Percent white alone
train_data |>
  ggplot(aes(x = non_return_rate, y = pct_NH_White_alone_CEN_2010)) +
  geom_point(alpha = 0.25) +
  theme_minimal()

# Percent below poverty level
train_data |>
  ggplot(aes(x = non_return_rate, y = pct_Prs_Blw_Pov_Lev_ACS_13_17)) +
  geom_point(alpha = 0.25) +
  theme_minimal()
```

### Transformations of non_return_rate

Here, we explore natural logarithm and logit transformations of non_return_rate.

```{r}
train_data |>
  mutate(
    raw_non   = non_return_rate,
    log_non = log(non_return_rate + 0.001),
    logit_non = log(non_return_rate/ (1 - non_return_rate + 0.001))) |>
  pivot_longer(c(raw_non, log_non, logit_non)) |>
  ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~name, scales = "free") +
    labs(title = "Non Return Rate Transformations") +
    theme_minimal()
```

There are also other variables, in particular the income, home value, and population variables which could benefit from a log.

The population variables are: Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17, Pop_under_5_CEN_2010, Pop_18_24_CEN_2010 Value: Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17 Otherss: pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17, pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17, pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010, pct_No_Plumb_ACS_13_17

### Correlations

Target Correlation: Typically you better features (non-target variables) that are highly correlated with target variables as they provide more predictive power.

```{r}
numeric_vars <- train_data |>
  select(where(is.numeric), -non_return_rate)

# find correlations
correlations <- cor(numeric_vars, train_data$non_return_rate, use = "complete.obs") 
  
# Convert to dataframe and sort
non_cor <- data_frame(
  variable = rownames(correlations),
  correlation = (correlations)) |>
  arrange(desc(abs(correlation)))

non_cor
```

One potential issue is a lot of these variables measure the same thing just in different survey (pct_NH_White_alone_CEN_2010, pct_NH_White_alone_ACS_13_17)

Feature Correlation: One potential issue can be when features are highly correlated with one another. As one example, the percentage of families that speak Tagalog and Spanish may both contribute similar information to our model. The inclusion of both can lead to overfitting and make our estimates less stable. This issue doesn't matter as much for regularized models but does for regular regression.

Here, we examine correlation clusters.

```{r}
corr_mat <- cor(numeric_vars, use = "pairwise.complete.obs")

# distance based
dist_mat <- as.dist(1 - abs(corr_mat))

# hierarchical clustering
hc <- hclust(dist_mat, method = "average")

plot(hc, labels = FALSE, main = "Correlation Clusters")

# Id'ing cluster
clusters <- cutree(hc, h = 0.15)   # 1 - .85
table(clusters)
```

There are way too many correlated features to even begin to count. We approach this issue by creating correlation clusters, and then keep one representative variable per each cluster to mitigate instances of high collinearity within predictor variables. We then create a list of “redundant variables” (vars_to_drop) that represent any excess variables within clusters. This cuts our number of predictors by roughly 100, and decreases the computational complexity of the model.

```{r}
# Now we'll attach cluster numbers to each variable
cluster_df <- data.frame(
  variable = names(numeric_vars),
  cluster  = clusters)

representatives <- cluster_df |>
  group_by(cluster) |>
  slice(1) |> 
  pull(variable)

vars_to_drop <- setdiff(cluster_df$variable, representatives)

vars_to_drop
```

## Set up cross-validation and parallelization

```{r}
# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)
```

```{r}
# Set up parallelization
library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores() - 1)
registerDoParallel(cl)
```

# Predictive Modeling

## Model 1: Parametric

### Plain Vanilla Linear Regression

First, we try a "plain vanilla" linear regression model without removing the redundant variables we identified earlier. We remove character variables, variables that are similarly defined as our outcome variable, and a variable with a very high level of missingness. We use median imputation for the rest of the variables with missingness.

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  # Specify and remove id variables 
  update_role(State_name, County_name,
              new_role = "id") |>
  # Remove other variables 
  step_rm(has_role("id"),
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          # Remove variables similar to the outcome variable
          Self_Response_Rate_ACS_13_17,
          Low_Response_Score,
          # Remove variable with high missingness
          pct_TEA_Update_Leave_CEN_2010) |> 
  # Median imputation for other variables with missingness 
  step_impute_median(all_numeric_predictors())

prep(model1_rec) |> juice() |> names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()
```

Next, we extract the 20 most important predictor variables from our linear regression model.

```{r}
# Variable importance
model0_final <- model0_wf |>
  last_fit(split) 

model0_final |>
  extract_fit_parsnip() |>
  vip(num_features = 20) 

#Extracting the most important for later use 
fit <- model0_final |> 
  extract_fit_parsnip()

imp_tbl <- vi(fit)

# sort by absolute importance
imp_tbl <- imp_tbl |> arrange(desc(abs(Importance)))

# convert to a vector of variable names
top_vars <- imp_tbl$Variable[1:20]

top_vars

top_vars_clean <- intersect(top_vars, names(train_data))
```

### Linear regression, removing redundant features

Then, we try another linear regression model, this time we remove the redundant variables we identified earlier. Same as the "plain vanilla" linear regression, we remove character variables, variables that are similarly defined as our outcome variable, and a variable with a very high level of missingness. We use median imputation for the rest of the variables with missingness.

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
    # Specify and remove id variables 
  update_role(State_name, County_name,
              new_role = "id") |>
  # Remove other variables 
  step_rm(has_role("id"),
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          # Remove variables similar to the outcome variable
          Self_Response_Rate_ACS_13_17,
          Low_Response_Score) |>
  # Remove redundant/highly correlated predictors 
  step_rm(all_of(vars_to_drop)) |> 
  # Median imputation for other variables with missingness 
  step_impute_median(all_numeric_predictors())

prep(model1_rec) |> 
  juice() |> 
  names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()
```

Removing redundant variables did not change the RMSE.

### Linear regression model with transformations

Next, we try a third linear regression model, with the only change from the previous model being that we apply natural logarithmic transformations of the following variables:

-   Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17

-   Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17

-   pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17, pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17, pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010, pct_No_Plumb_ACS_13_17

We also normalize all predictors.

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  # Specify and remove id variables 
  update_role(State_name, County_name,
              new_role = "id") |>
  # Remove other variables
  step_rm(has_role("id"),
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          # Remove variables similar to the outcome variable
          Self_Response_Rate_ACS_13_17,
          Low_Response_Score) |> 
  # Apply log transformations 
  step_log(Tot_Population_ACS_13_17, Males_ACS_13_17, Females_ACS_13_17, offset = 1) |>
  step_log(Med_HHD_Inc_ACS_13_17, Med_House_Value_ACS_13_17, offset = 1) |>
  step_log(pct_Age5p_French_ACS_13_17, pct_College_ACS_13_17,
           pct_Prs_Blw_Pov_Lev_ACS_13_17, pct_TwoPHealthIns_ACS_13_17,
           pct_Female_No_HB_CEN_2010, pct_Sngl_Prns_HHD_CEN_2010,
           pct_No_Plumb_ACS_13_17, offset = 1e-6) |>
  # Remove redundant/highly correlated predictors 
  step_rm(all_of(vars_to_drop)) |> 
  # Impute with median for numeric predictors' missing values
  step_impute_median(all_numeric_predictors()) |>
  # Normalize all predictors 
  step_normalize(all_numeric_predictors()) 

prep(model1_rec) |> juice() |> names()

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

model0_resamples <- model0_wf |>
  fit_resamples(resamples = folds)

model0_resamples |>
  collect_metrics()

model0_final <- model0_wf |>
  last_fit(split) |>
  collect_metrics()
```

Applying log transformations did not improve the RMSE.

### Linear Regression with PCA hyperparameter tuning

Given the large number of variables in this dataset, we apply a principal component analysis (PCA) tuning grid. PCA can be especially helpful for this task because it handles multicollinearity, reduces overfitting, and reduces the number of predictors, which saves computational expense. Our tuning grid tunes over a range of 50 to 150 principal components.

```{r}
set.seed(20260412)

# Recipe
model1_rec <- recipe(non_return_rate ~ ., train_data) |>
  # Specify and remove id variables
  update_role(State_name, County_name,
              new_role = "id") |>
  step_rm(has_role("id"),
          # Remove variable with high missingness
          pct_TEA_Update_Leave_CEN_2010,
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          # Remove variables similar to the outcome variable 
          Self_Response_Rate_ACS_13_17,
          Low_Response_Score) |> 
  # Impute median for numeric predictors 
  step_impute_median(all_numeric_predictors()) |>
  # Normalize all predictors
  step_normalize(all_numeric_predictors()) |>
  # Apply PCA hyperparameter tuning 
  step_pca(all_numeric_predictors(), num_comp = tune::tune())

model0_mod <- linear_reg() |>
  set_mode(mode = "regression") |>
  set_engine(engine = "lm")

model0_wf <- workflow() |>
  add_recipe(model1_rec) |>
  add_model(model0_mod)

# Add PCA grid
pca_grid <- grid_regular(
  num_comp(range = c(50, 150)),
  levels = 10)

model0_resamples <- tune_grid(
  model0_wf,
  resamples = folds,
  grid = pca_grid)

model0_resamples |>
  collect_metrics()
```

Our RMSE remains high for the levels of PCA tried in the grid.

### Elastic Net

Next, we try applying elastic net regularization with hyperparameter tuning for both the penalty and the mixture. Our motivation is to try to improve the linear regression model's generalization to unseen data.

```{r}
set.seed(20260412)

# Recipe
en_rec <- recipe(non_return_rate ~ ., train_data) |>
  # Specify and remove id variables
  update_role(State_name, County_name,
              new_role = "id") |>
  step_rm(has_role("id"),
          # Remove variable with high missingness
          pct_TEA_Update_Leave_CEN_2010,
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          # Remove variables similar to the outcome variable 
          Self_Response_Rate_ACS_13_17,
          Low_Response_Score) |> 
  # Impute median for numeric predictors 
  step_impute_median(all_numeric_predictors()) |>
  # Normalize all predictors
  step_normalize(all_numeric_predictors())

# Mode
en_mode <- linear_reg(penalty = tune(), mixture = tune()) |> # works best for many variables 
  set_mode(mode = "regression") |>
  set_engine(engine = "glmnet") 

# Workflow
en_wf <- workflow() |>
  add_recipe(en_rec) |>
  add_model(en_mode)

# Resamples
en_grid <- grid_regular(
  penalty(), 
  mixture(),
  levels = 20
)

# Tuning grid
en_tuning <- tune_grid(
  en_wf,
  resamples = folds,
  grid = en_grid
)

show_best(en_tuning)

# Finalize workflow
en_best <- select_best(en_tuning, metric = "rmse")

en_final_wf <- finalize_workflow(
  en_wf,
  en_best)
```

We do see a slight improvement in our RMSE by using elastic net regression. Next, we extract the top 20 most important predictors.

```{r}
# Variable importance
en_fit <- fit(en_final_wf, data = train_data)

top_50_vars <- en_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 50) 

top_50_vars
```

All our linear regression models employing feature engineering, selection, and regularization had similar RMSE outcomes. Next, to further improve RMSE, we adopt a more complex model that will be better at learning the non-linear patterns in the data.

## Model 2: Regression tree (Non-parametric)

For our second type of model, we have selected a regression tree, a non-parametric approach. We tune the following three regression tree hyperparameters:

-   Cost complexity - balances accuracy with tree size

-   Tree depth - the length from the root node to the farthest leaf

-   min_n - the minimum number of data points a node must contain before further splitting

After removing "id" variables, character variables, and variables highly similar to the outcome variable, we select only the top 50 most important features from elastic net regression to reduce computational expense.

```{r}
set.seed(20260414)
tic()
# Recipe
model2_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  # Specify and remove id variables
  update_role(GIDTR, State_name, County_name, new_role = "id") |>
  # Remove highly redundant/correlated variables
  step_rm(all_of(vars_to_drop)) |> 
  step_rm(has_role("id"),
          # Remove character variables 
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, 
          # Remove variables similar to the outcome variable 
          Low_Response_Score,
          Self_Response_Rate_ACS_13_17) |> 
  # Select the top 50 most important features from elastic net
  step_select(top_50_vars) |>
  # Impute median for missingness 
  step_impute_median(all_numeric_predictors())

# Model
model2_mod <- decision_tree(
    cost_complexity = tune(),
    tree_depth = tune(),
    min_n = tune()
  ) |>
  set_engine(engine = "rpart") |>
  set_mode(mode = "regression")

# Tuning grid
model2_grid <- grid_space_filling(
  cost_complexity(range = c(-5, -1)), 
  tree_depth(range = c(3, 15)), 
  min_n(range = c(2, 50)),
  size = 20
)

# Workflow
model2_wf <-
  workflow() |>
  add_recipe(recipe = model2_rec) |>
  add_model(spec = model2_mod)

# Resamples
model2_resamples <- model2_wf |>
  tune_grid(resamples = folds,
            grid = model2_grid,
            metrics = metric_set(rmse),
            control = control_grid(parallel_over = "everything"))

# Collect metrics
collect_metrics(model2_resamples)
toc()
```

```{r}
model2_resamples |>
  show_best()

# Visualize the results
autoplot(model2_resamples)
```

In the top results, the tree depth varies widely, the cost complexity is very small, and the min_n is moderately large. These results suggest that the tree depth doesn't matter very much, but smaller cost complexity and splitting on moderate/large nodes are best.

\*Note to Miguel: I deleted the version of regression tree with a very long run time (12666.4 sec). In this now deleted version, I tried increasing the size of the grid to 100 to try more options, a different range for cost complexity, and a larger range of min_n. I also tried changing the order of my recipe steps to remove variables with low variance before imputation, and the threshold of step_corr to 0.75 instead of 0.9 since we already removed variables with high correlation. It had very little effect on RMSE. I welcome any changes to feature selection/engineering you want to add to improve Model 2, or just leave it as is.
