---
title: "Model 3 Ensembling"
author: "Sarah Krause and Miguel Villa"
format: html
editor: visual
embed-resources: true
execute:
  cache: true
---
## Repeat of Earlier Steps
```{r message=FALSE, warning=FALSE}
#| code-fold: true

options(scipen = 999)

library(tidyverse)
library(here)
library(tidymodels)
library(skimr)
library(janitor)
library(corrplot)
library(tictoc)
library(vip)

data <- read.csv(here("data/assignment_modeling.csv"))

## Split the data

set.seed(20260412)
# Place 80% of the data in the training data
split <- initial_split(data, prop = 0.8)

# Obtain the training data
train_data <- training(split)

## Set up cross-validation

# 5 fold cross-validation
folds <- vfold_cv(train_data, v = 5)

# specifying only numeric variables 
numeric_vars <- train_data |>
  select(where(is.numeric), -non_return_rate)

# find correlations
correlations <- cor(numeric_vars, train_data$non_return_rate, use = "complete.obs") 
  
# Convert to dataframe and sort
non_cor <- tibble(
  variable = rownames(correlations),
  correlation = (correlations)) |>
  arrange(desc(abs(correlation)))
corr_mat <- cor(numeric_vars, use = "pairwise.complete.obs")

# distance based
dist_mat <- as.dist(1 - abs(corr_mat))

# hierarchical clustering
hc <- hclust(dist_mat, method = "average")

# Id'ing cluster
clusters <- cutree(hc, h = 0.15)   # 1 - .85

cluster_df <- data.frame(
  variable = names(numeric_vars),
  cluster  = clusters)

representatives <- cluster_df |>
  group_by(cluster) |>
  slice(1) |> 
  pull(variable)

vars_to_drop <- setdiff(cluster_df$variable, representatives)
```


### Random Forrest Model 1
For our third model, we have will first try random forest, an ensembling approach. We tune the hyperparameters *min_n*, the minimum number of data points required in a node before further splitting, and *mtry*, the number of predictors considered at each split using the latin hypercube tuning grid to save computation time. For *min_n*, we tune over a range of 5 to 40 data points required in a node before further splitting. For *mtry,* we tune over a range of 5 to 50 predictors to be considered at each split. During tuning, we use 100 trees. The number of trees will be increased to between 200-500 for final model fitting. We also use parallelization to reduce computation time.

In the recipe, several variables are removed, as well as variables with low variance and high correlation with each other. These steps reduce redundancy and also save computation time. We also use 3-fold cross-validation instead of 5-fold cross-validation.

```{r message=FALSE, warning=FALSE}
#| freeze: true

set.seed(20260412)
tic()

# Recipe
model3_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  update_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"),
          Aggr_House_Value_ACS_13_17, 
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17, # remove the problematic fact 
          Low_Response_Score,
          Self_Response_Rate_ACS_13_17) |> 
  step_rm(all_of(vars_to_drop)) |> # removing highly correlated vars
  step_impute_median(all_numeric_predictors()) 

# Model
model3_mod <- rand_forest(min_n = tune(), # hyperparameter tuning
                          mtry = tune(),
                          trees = 100) |>
  set_engine(engine = "ranger",
             importance = "impurity",
             num.threads = parallel::detectCores() - 1) |> # parallelization
  set_mode(mode = "regression")

# Workflow
model3_wf <- workflow() |>
  add_recipe(model3_rec) |>
  add_model(model3_mod)

# Tuning grid
model3_grid <- grid_space_filling(
  mtry(range = c(5, 50)),
  min_n(range = c(5, 40)),
  size = 20
)

# Resamples
model3_resamples <- tune_grid(
  model3_wf,
  resamples = folds,
  grid = model3_grid,
  control = control_grid(parallel_over = "everything")
)

# Collect metrics
collect_metrics(model3_resamples)

toc()

# Best parameters
best_rf <- select_best(model3_resamples, metric = "rmse")

# Finalize and fit 
final_rf_wf <- finalize_workflow(model3_wf, best_rf)

final_rf_fit <- last_fit(final_rf_wf, split = split)



```

```{r}
# Calculate fit
collect_metrics(final_rf_fit)
```

### XGB Model
An XG Boost model is our final model and is consider for it's considerable benefits in terms of bias reduction as a model that sequential fits predictive models on the previous iteration of the sequence. Typically this means the model is fit on the residual of the previous models. 

```{r message=FALSE, warning=FALSE}
#| freeze: true

set.seed(20260412)
tic()

library(xgboost)
library(doParallel)

# Parallelize across all available cores
registerDoParallel(cores = parallel::detectCores() - 1)

# Recipe 
model5_rec <- recipe(non_return_rate ~ ., data = train_data) |>
  update_role(GIDTR, State_name, County_name, new_role = "id") |>
  step_rm(has_role("id"),
          Aggr_House_Value_ACS_13_17, 
          Aggregate_HH_INC_ACS_13_17,
          avg_Agg_HH_INC_ACS_13_17,
          avg_Agg_House_Value_ACS_13_17,
          Low_Response_Score,
          Self_Response_Rate_ACS_13_17) |> 
  step_rm(all_of(vars_to_drop)) |>
  step_impute_median(all_numeric_predictors()) 

prep(model5_rec) |> juice() |> names()

xgb_spec <- boost_tree(
  trees = 500,
  stop_iter = 20,
  tree_depth = tune(), min_n = tune(),
  loss_reduction = tune(),
  sample_size = tune(), mtry = tune(),
  learn_rate = tune()) |>
  set_engine("xgboost", 
             validation = 0.1,      
             nthread = 1) |>       
  set_mode("regression")

xgb_grid <- grid_latin_hypercube(
  tree_depth(range = c(3, 8)),       
  min_n(range = c(2, 10)),
  loss_reduction(range = c(-3, -1)),
  sample_size = sample_prop(range = c(0.6, 0.9)), 
  mtry(range = c(5, 20)),           
  learn_rate(range = c(-3, -1)),    
  size = 15
)

xgb_wf <- workflow() |>
  add_recipe(model5_rec) |>
  add_model(xgb_spec)

xgb_resamples <- tune_grid(
  xgb_wf,
  resamples = folds,
  grid = xgb_grid,
  control = control_grid(
    parallel_over = "everything",
    save_pred = FALSE,         
    verbose = TRUE      
  )
)

collect_metrics(xgb_resamples)

# Best parameters
best_xgb <- select_best(xgb_resamples, metric = "rmse")

final_xgb_wf <- finalize_workflow(xgb_wf, best_xgb)

# fit on full split
final_xgb_fit <- last_fit(final_xgb_wf, split = split)

# calculate fit
collect_metrics(final_xgb_fit)

toc()

```


## Fitting the xgb (best) model on the implementation data
```{r}
# Load the implementation file
implementation_data <- read_csv(here("data/assignment_implementation.csv")) |>
  mutate(GIDTR = as.numeric(GIDTR))

# extract workflow
final_xgb_model <- extract_workflow(final_xgb_fit)

# Predictions
predictions <- predict(final_xgb_model, new_data = implementation_data)

# data frame with performance
performance <- implementation_data |>
  select(GIDTR) |>
  bind_cols(predictions)

# Creating CSV with just id and predictions
write_csv(performance, "best_model.csv")
```